---
title: "class 15"
author: "Tingyi Yang"
date: "2/26/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##use source("http://bioconductor.org/biocLite.R")
##    biocLite("DESeq2") to install bioconductor
```{r}
library(BiocInstaller)
```

```{r}
counts <- read.csv("data/airway_scaledcounts.csv", stringsAsFactors = FALSE)
metadata <-  read.csv("data/airway_metadata.csv", stringsAsFactors = FALSE)
```
```{r}
head(counts)
```

```{r}
head(metadata)
```
```{r}
View(metadata)

```

```{r}
control <- metadata[metadata[,"dex"]=="control",]
control.mean <- rowSums( counts[ ,control$id] )/4 
names(control.mean) <- counts$ensgene
```

```{r}
colnames(counts)
```

```{r}
metadata$id
```

Check for correspondence for metadata and countdata
```{r}
all(colnames(counts)[-1] == metadata$id)
##all() checks everything inside is true
```

TASK: 1. Find which columns in **counts** corresponds to the control conditions(i.e. no drug!). Then we can calculate the mean count value for each gene.

```{r}
control.inds <- metadata$dex == "control"
control.ids <- metadata$id[control.inds]

head(counts[, control.ids])
```
```{r}
rowSums(counts[,control.ids])/ncol(counts[,control.ids])

```

add mean counts value for all genes:
```{r}
control.mean <- rowSums(counts[, control.ids])/ncol(counts[, control.ids])
names(control.mean) <- counts$ensgene
head(control.mean)
```

Now do the same thing with the treated samples. (i.e. **treated** columns)
```{r}
treated.inds <- metadata$dex == "treated"
treated.ids <- metadata$id[treated.inds]

head(counts[, treated.ids])
```

```{r}
rowSums(counts[,treated.ids])/ncol(counts[,treated.ids])
```

```{r}
treated.mean <- rowSums(counts[, treated.ids])/ncol(counts[, treated.ids])
names(treated.mean) <- counts$ensgene
head(treated.mean)
```


```{r}
meancounts <- data.frame(control.mean, treated.mean)
```
```{r}
colSums(meancounts)
```

Now we can plot the drug treated vs non-drug control count values
```{r}
plot(meancounts$control.mean,meancounts$treated.mean)
```

Change this graph into a log plot since the plot right now is very clustered


```{r}
plot(log(meancounts$control.mean),log(meancounts$treated.mean))
```

 We usually look at the log2 of the fold change, because this has better mathematical properties.
```{r}
meancounts$log2fc <- log2(meancounts[,"treated.mean"]/meancounts[,"control.mean"])
head(meancounts)
```
 
```{r}
toy <- c(1,2,0,4,0)
toy == 0
which(toy == 0)
```
```{r}
toyy <- meancounts[1:6,]
toyy[,1:2]
which(toyy[,1:2] == 0, arr.ind = TRUE)
unique(which(toyy[,1:2] == 0, arr.ind = TRUE)[,1])
```
Remove the zero entry genes from meancounts to produce **mycounts**
```{r}
zero.vals <- which(meancounts[,1:2]==0, arr.ind=TRUE)

to.rm <- unique(zero.vals[,1])
mycounts <- meancounts[-to.rm,]
head(mycounts)
```

```{r}
up.ind <- mycounts$log2fc > 2
down.ind <- mycounts$log2fc < (-2)

```
```{r}
sum(up.ind)
```
```{r}
sum(down.ind)
```
There are `sum(up.ind)` up regulated genes and `sum(up.ind)` down regulated genes.

```{r}
head(mycounts[up.ind,])
```

Adding annotation:

```{r}
anno <- read.csv("data/annotables_grch38.csv")
head(anno)
```

Use **merge()** function to add annotation data from ensembl
```{r}
mycounts.annotated <- merge(x=mycounts,y=anno, by.x="row.names", by.y="ensgene")
head(mycounts.annotated)
```

## DESeq2 analysis 
now we do this analysis with p-values
```{r}
library("DESeq2")
```
```{r}
citation("DESeq2")
```

```{r}
dds <- DESeqDataSetFromMatrix(countData=counts, 
                              colData=metadata, 
                              design=~dex, 
                              tidy=TRUE)
dds

```
```{r}
dds <- DESeq(dds)
```

```{r}
res <- results(dds)
res
```
```{r}
summary(res)
```
```{r}
res01 <- results(dds, alpha=0.01)
summary(res01)
```

```{r}
plot(res$log2FoldChange, -log(res$padj))
```

Setup point color vector
```{r}
mycols <- rep("gray", nrow(res01))
mycols[abs(res01$log2FoldChange)>2] <- "red"
```
```{r}
##(inds--limit, True only)
inds<- (res01$padj<0.01)&(abs(res05$log2FoldChange)>2)
mycols[inds] <- "blue"
```
##Volcano plot with custom colors
```{r}
plot(res01$log2FoldChange, -log(res01$padj), col=mycols, ylab= "-Log(P-value)", xlab = "Log2(FoldChange)")

abline(v=c(-2,2), col="gray", lty=2)
abline(h=-log(0.1), col="gray", lty=2)
```


Dr. Grant's link:
http://tinyurl.com/class15-classmd 





































